---
output:
  pdf_document: default
  html_document: default
---
Red wine quality dataset exploration and summary by Mohamed Zeghlache
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library("PerformanceAnalytics")
```

```{r echo=FALSE, Load_the_Data}
# setting my working directory
setwd('C:/Users/Mohamed/Desktop/R Project')

# Loading the red wine Data set
wine_dataset <- read.csv('wineQualityReds.csv')
```

> **General information about the dataset**: 
This tidy data set contains 1,599 red wines with 11 variables on the chemical properties of the wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).
The classes are ordered and not balanced (e.g. there are munch more normal wines than excellent or poor ones). Outlier detection algorithms could be used to detect the few excellent or poor wines. Also, we are not sure if all input variables are relevant. So it could be interesting to test feature selection methods. 
Number of Instances: 1599.
Number of Attributes: 11 + output attribute

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
# displaying the structure of the dataset
str(wine_dataset)
```

as we can see, this dataset contains 1599 observations(rows) and 13 variables(attributes)
but the first variable is just an index and we don't need it in our analysis,so i'll delete it from the dataset.but the rest of the variables are just as described in the introduction.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# deleting the first variable X
wine_dataset[1] <- NULL

# display the structure of the dataset again
str(wine_dataset)
```

nice, now the dataset looks better.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# view a statistical summary on the dataset
summary(wine_dataset)
```

from this result, i can already see that some of these variables have outliers, but we'll get a closer look on that using plots.

## ploting

### fixed.acidity
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of fixed.acidity
p1 <- ggplot(aes(x = fixed.acidity) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = fixed.acidity) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```

these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right.so, i applied a log scale to the x axis to make the distribution look more normal(the one in the bottom).


### volatile.acidity
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of volatile.acidity
p1 <- ggplot(aes(x = volatile.acidity) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = volatile.acidity) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```

these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right.so, i applied a log scale to the x axis to make the distribution look more normal(the one in the bottom).


### citric.acid
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of citric.acid
p1 <- ggplot(aes(x = citric.acid) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = citric.acid) , data = wine_dataset)+
  geom_histogram()+
  scale_x_sqrt()

grid.arrange(p1,p2)
```

these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right.so, i applied a square root scale to the x axis to make the distribution look more normal(the one in the bottom).


### residual.sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of residual.sugar
p1 <- ggplot(aes(x = residual.sugar) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = residual.sugar) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```

these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right and has outliers.so, i applied a log10 scale to the x axis to make the distribution look more normal(the one in the bottom).

### chlorides
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of chlorides
p1 <- ggplot(aes(x = chlorides) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = chlorides) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```


these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right and has outliers.so, i applied a log10 scale to the x axis to make the distribution look normal(the one in the bottom).


### free.sulfur.dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of free.sulfur.dioxide
p1 <- ggplot(aes(x = free.sulfur.dioxide) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = free.sulfur.dioxide) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```


these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right and has outliers.so, i applied a log10 scale to the x axis to make the distribution look more normal or bimodal(the one in the bottom).

### total.sulfur.dioxide
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of total.sulfur.dioxide
p1 <- ggplot(aes(x = total.sulfur.dioxide) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = total.sulfur.dioxide) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2)
```


these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right and has outliers.so, i applied a log10 scale to the x axis to make the distribution look normal (the one in the bottom).

### density
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of density
ggplot(aes(x = density) , data = wine_dataset)+
  geom_histogram()
```


this histogram looks perfectly normaly ditributed.

### pH
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of pH
ggplot(aes(x = pH) , data = wine_dataset)+
  geom_histogram()
```

this histogram also looks very normaly ditributed, but it might contain some outliers.

### sulphates
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of sulphates
p1 <- ggplot(aes(x = sulphates) , data = wine_dataset)+
  geom_histogram()

p2 <- ggplot(aes(x = sulphates) , data = wine_dataset)+
  geom_histogram()+
  scale_x_log10()

grid.arrange(p1,p2,ncol=1)
```

these two histograms are both for the same variable,the histogram that's in the top looks skewed to the right and has outliers.so, i applied a log10 scale to the x axis to make the distribution look normal (the one in the bottom).


### alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a histogram of alcohol
ggplot(aes(x = alcohol) , data = wine_dataset)+
  geom_histogram()
```

this histogram is skewed to the right.

### quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot a bar plot of quality
ggplot(aes(x = quality) , data = wine_dataset)+
  geom_bar()
```

this bar plot shows the distribution of the quality values, it looks like the quality of the red wine is between 3 and 8.with most of the red wine having a quality of 5 or 6.




# Univariate Analysis


### What is the structure of your dataset?

the red wine dataset contains 12 attributes(variables), 11 of them are numerical values for chemical tests made on the wine, and the last attribute is the output of those chemical tests(the quality). and it consists of 1599 observations.


### What is/are the main feature(s) of interest in your dataset?

clearly, the main feature of interest in the dataset is the quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

all the other 11 features may have an effect over the quality because the quality depends on them. although, they might have different levels of correlation to the quality.

### Did you create any new variables from existing variables in the dataset?

No, i did not.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

There were no unusual distributions.
i did not perform any operations to tidy the data, it's already tidy. I've just dropped a column because it represented the index of the rows which is "redundant".

# Bivariate Plots Section

we'll first start by drawing a correlation chart of all the variables
```{r echo=FALSE, Bivariate_Plots}
# drawing a correlation chart
chart.Correlation(wine_dataset)
```

on this plot we can see the distributions of all the variables on the diagonal, and we can also see the scatter plots of different variable below the diagram, and on top of the diagram we can see the significance and correletion between the variables.

based on the scatter plots and the correlation test scores; i'll perform some biavariate plots.
i will not draw many plots, because this chart has it all.
i'll draw two plots of the two most correlated variables to the output variable depending on this chart, and i'll draw a couple more plots of other variables that 


### volatile.acidity vs quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = quality) , data = wine_dataset)+
  geom_point()+
  scale_x_continuous(limits = c(0.1,1.4))+
  geom_smooth()
```

we can see that there is a small correlation between the volatile.acidity and the quality, as the volatile.acidity increases the quality decreases.with only on outlier that i have eliminated from the plot.


### alcohol vs quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = quality) , data = wine_dataset)+
  geom_point()+
  scale_x_continuous(limits = c(8.5,14.5))+
  geom_smooth()
```

we can see that there is a small correlation between the alcohol and the quality, as the alcohol increases the quality also increases, with only on outlier that i have eliminated from the plot.

### fixed.acidity vs density

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = density) , data = wine_dataset)+
  geom_point()+
  geom_smooth()
```

we can also notice that there is a strong correlation between the fixed.acidity and the density of the wine.


### fixed.acidity vs pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = pH) , data = wine_dataset)+
  geom_point()+
  geom_smooth()
```

we can also notice that there is a strong negative correlation between the fixed.acidity and the pH of the wine.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

according to the correlation chart, the strongest relationships for the feaute of interest(quality) are with the alcohol and the citric.acid

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
yeah, there were some strong correlation test scores between other features as (fixed.acidity vs pH) and (fixed.acidity vs density)

### What was the strongest relationship you found?
the strongest relationship i found was (fixed.acidity vs pH)



# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = fixed.acidity , y = pH) , 
       data = subset(wine_dataset, quality > 6))+
  geom_jitter(aes(color = quality) , alpha = 0.5)
```

i can't see a clear pattern for the high quality wine's relationship to pH and fixed.acidity.



```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity , y = density) , 
       data = subset(wine_dataset, quality > 6))+
  geom_jitter(aes(color = quality) , alpha = 0.5)
```

i can't see a clear pattern for the high quality wine's relationship to pH and fixed.acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid , y = alcohol) ,
       data = wine_dataset)+
  geom_jitter(aes(color = quality) , alpha = 0.5)+
  scale_x_continuous(limits = c(0,0.75))
```

the increase of both alcohol and citric acid results in an increase in th quality of red wine.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates , y = alcohol) ,
       data = wine_dataset)+
  geom_jitter(aes(color = quality) , alpha = 0.25)+
  scale_x_continuous(limits = c(0.3,1.3))
```

a high quality of alcohol has a value of sulphates between (0.65 and 1.25) and a value of alcohol between (5 and 13) "on average".


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

yeah, i think that the citric acid and the alcohol stregthen each other, and alcohol and sulphates also stregthen each other.

### Were there any interesting or surprising interactions between features?
i think that alcohol has a big impact on almost all of the other variables .

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}
# plot a bar plot of quality
ggplot(aes(x = quality ) , data = wine_dataset)+
  geom_bar(aes(fill = I('orange')))+
  scale_x_continuous(breaks = seq(0,8,1))+
  ggtitle('the distribution of qualities over the dataset')
```



### Description One
this bar plot shows that the quality of the red wine ranges between 3 and 8, but most the red wine in the dataset have a quality of 5 or 6.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = fixed.acidity, y = density, color = 'red') , data = wine_dataset)+
  geom_point()+
  scale_x_continuous(limits = c(4,14))+
  ggtitle('correlation between density and fixed_acidity')
```


### Description Two
this plot is interesting because we can see how the density and the fixed.acidity strengthen each other. The plot shows a strong positive correlation.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = alcohol , y = citric.acid) ,
       data = wine_dataset)+
  geom_point(aes(color = quality) , alpha = 0.5, position = 'jitter')+
  scale_x_continuous(limits = c(9,14))+
  scale_y_continuous(limits = c(0,0.8))+
  ggtitle('correlation between citric_acid and alcohol')
```



### Description Three
in this plot we can see that the increase of both alcohol and citric acid results in an increase in the quality of red wine.
------

# Reflection

this was a very interesting dataset with many variables to explore.

one of the struggles that i went through is deciding if any variable really had a significant effect on the quality of the wine, because the dataset didn't have many high quality wine samples(unbalanced).

for me, drawing the bar chart had a very usefel impact on defining a path of the exploration of this dataset.

i think that a future work that could be done with the dataset is to give all of the wine samples the same alcohol value and see if the distribution of the quality in the dataset will have a significant change.

